Analysis-ready VCF at Biobank scale using Zarr

### 2025-01-30 paper.tex
Fix spruce m number. 3.75 billion!

To demonstrate the versatility of \texttt{vcf2zarr}, in this section
we include a case study of a dataset originating from a species whose
genome properties differ substantially from human. The conifer Norway
spruce, \emph{Picea abies}, is one of the largest and most
ecologically important species on Earth and is distributed over large
parts of the Northern hemisphere. One of its features is its large
genome, consisting of 12 autosomes with a genome size of 19.6
Gb~\cite{nystedt_NorwaySpruceGenome_2013}. Here, we assess the
performance of \texttt{vcf2zarr} as applied to 1,063 whole-genome
resequenced spruce individuals from a recent study (in preparation),
where resequencing data has been mapped to a chromosome-scale
reference genome. The dataset consists of approximately 3.8 million
single-nucleotide variants and small indels from the 12 autosomes
totalling 7.4TiB of VCF data after \texttt{bgzip} compression,
distributed over 165 VCF files. 
% Since the dataset originally was
% generated for downstream analyses that require genotype likelihoods,
% the genotypes include the PL field.

@@

\begin{table}
\caption{Summary for VCF Zarr conversion of the Norway Spruce
%TODO The variants number here does not match the text "3.8 million"
WGS data on all 12 autosomes (1,063 samples, 374,517,0452 variants)
consisting of 34 fields and 2.3 TiB of storage ($\sim 3.2$X smaller 
than source gzipped VCFs). The chunk size is 10,000 variants $\times$
1,063 samples (i.e., a single sample chunk). This uses local 
alleles to reduce the 
space required by the PL field; see the text for discussion.
Shown are the top 21 fields consuming at least 0.01\% of the 
overall storage (see Table~\ref{tab-genomics-england-data} for column details).
\label{tab-spruce-data}}
\begin{tabular}{llS[table-format=3.2]S[table-format=3.2]S[table-format=3.2]}
\toprule
{Field} & {Type} & {Storage} & {Compress} & {\% Total} \\
\midrule
call\_LPL & int16 & 1.26 TiB & 17.0 & 54.76\% \\
call\_LA & int8 & 415.39 GiB & 18.0 & 17.63\% \\
call\_genotype & int8 & 381.28 GiB & 19.0 & 16.18\% \\
call\_genotype\_mask & bool & 103.84 GiB & 71.0 & 4.41\% \\
variant\_DP4 & int32 & 26.24 GiB & 2.1 & 1.11\% \\
variant\_MQSBZ & float32 & 13.08 GiB & 1.1 & 0.56\% \\
variant\_RPBZ & float32 & 13.04 GiB & 1.1 & 0.55\% \\
variant\_MQBZ & float32 & 12.97 GiB & 1.1 & 0.55\% \\
variant\_SCBZ & float32 & 12.94 GiB & 1.1 & 0.55\% \\
variant\_quality & float32 & 12.84 GiB & 1.1 & 0.55\% \\
variant\_VDB & float32 & 12.84 GiB & 1.1 & 0.55\% \\
variant\_BQBZ & float32 & 12.64 GiB & 1.1 & 0.54\% \\
variant\_SGB & float32 & 12.12 GiB & 1.2 & 0.51\% \\
variant\_position & int32 & 9.65 GiB & 1.4 & 0.41\% \\
variant\_AC & int16 & 7.42 GiB & 2.8 & 0.31\% \\
variant\_DP & int32 & 6.71 GiB & 2.1 & 0.28\% \\
variant\_allele & str & 5.37 GiB & 21.0 & 0.23\% \\
variant\_AN & int16 & 3.05 GiB & 2.3 & 0.13\% \\
call\_genotype\_phased & bool & 1.75 GiB & 2100.0 & 0.07\% \\
variant\_filter & bool & 1.46 GiB & 2.4 & 0.06\% \\
variant\_MQ & int8 & 489.41 MiB & 7.3 & 0.02\% \\
\bottomrule
\end{tabular}
\end{table}

>>>>

To demonstrate the versatility of \texttt{vcf2zarr}, in this section
we include a case study of a dataset originating from a species whose
genome properties differ substantially from human. The conifer Norway
spruce, \emph{Picea abies}, is one of the largest and most
ecologically important species on Earth and is distributed over large
parts of the Northern hemisphere. One of its features is its large
genome, consisting of 12 autosomes with a genome size of 19.6
Gb~\cite{nystedt_NorwaySpruceGenome_2013}. Here, we assess the
performance of \texttt{vcf2zarr} as applied to 1,063 whole-genome
resequenced spruce individuals from a recent study (in preparation),
where resequencing data has been mapped to a chromosome-scale
reference genome. The dataset consists of approximately 3.75 billion
single-nucleotide variants and small indels from the 12 autosomes
totalling 7.4TiB of VCF data after \texttt{bgzip} compression,
distributed over 165 VCF files. 
% Since the dataset originally was
% generated for downstream analyses that require genotype likelihoods,
% the genotypes include the PL field.

@@

\begin{table}
\caption{Summary for VCF Zarr conversion of the Norway Spruce
WGS data on all 12 autosomes (1,063 samples, 
%3745170452
3,745,170,452 variants)
consisting of 34 fields and 2.3 TiB of storage ($\sim 3.2$X smaller 
than source gzipped VCFs). The chunk size is 10,000 variants $\times$
1,063 samples (i.e., a single sample chunk). This uses local 
alleles to reduce the 
space required by the PL field; see the text for discussion.
Shown are the top 21 fields consuming at least 0.01\% of the 
overall storage (see Table~\ref{tab-genomics-england-data} for column details).
\label{tab-spruce-data}}
\begin{tabular}{llS[table-format=3.2]S[table-format=3.2]S[table-format=3.2]}
\toprule
{Field} & {Type} & {Storage} & {Compress} & {\% Total} \\
\midrule
call\_LPL & int16 & 1.26 TiB & 17.0 & 54.76\% \\
call\_LA & int8 & 415.39 GiB & 18.0 & 17.63\% \\
call\_genotype & int8 & 381.28 GiB & 19.0 & 16.18\% \\
call\_genotype\_mask & bool & 103.84 GiB & 71.0 & 4.41\% \\
variant\_DP4 & int32 & 26.24 GiB & 2.1 & 1.11\% \\
variant\_MQSBZ & float32 & 13.08 GiB & 1.1 & 0.56\% \\
variant\_RPBZ & float32 & 13.04 GiB & 1.1 & 0.55\% \\
variant\_MQBZ & float32 & 12.97 GiB & 1.1 & 0.55\% \\
variant\_SCBZ & float32 & 12.94 GiB & 1.1 & 0.55\% \\
variant\_quality & float32 & 12.84 GiB & 1.1 & 0.55\% \\
variant\_VDB & float32 & 12.84 GiB & 1.1 & 0.55\% \\
variant\_BQBZ & float32 & 12.64 GiB & 1.1 & 0.54\% \\
variant\_SGB & float32 & 12.12 GiB & 1.2 & 0.51\% \\
variant\_position & int32 & 9.65 GiB & 1.4 & 0.41\% \\
variant\_AC & int16 & 7.42 GiB & 2.8 & 0.31\% \\
variant\_DP & int32 & 6.71 GiB & 2.1 & 0.28\% \\
variant\_allele & str & 5.37 GiB & 21.0 & 0.23\% \\
variant\_AN & int16 & 3.05 GiB & 2.3 & 0.13\% \\
call\_genotype\_phased & bool & 1.75 GiB & 2100.0 & 0.07\% \\
variant\_filter & bool & 1.46 GiB & 2.4 & 0.06\% \\
variant\_MQ & int8 & 489.41 MiB & 7.3 & 0.02\% \\
\bottomrule
\end{tabular}
\end{table}
